Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns

### 2019-08-06 ms/manuscript.Rmd
Fix some spelling errors

\clearpage
<!-- extra text-->
## Introduction

@@

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA). A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger. Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from July 18, 2012 to July 7, 2013 at six plots, and a final survey from July 7, 2013 to July 5, 2014 at 15 plots. Two plots were not included in the environmental survey. We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part of the final survey period (March 12--July 5, 2014), so we used data for the same dates from the preliminary survey period for this datalogger instead. Three other dataloggers also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis. Other days missing data for any datalogger during the final survey period were excluded from the dataset. The final dataset included `r total_days` days of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers. We calculated vapor pressure deficit (VPD) from relative humidity and temperature using the \package{plantecophys} R package [@Duursma2015].

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs \@ref(tab:phylosig-cont), \@ref(tab:phylosig-binary), \fig \@ref(fig:tree)), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \@ref(tab:phylosig-cont)). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \@ref(tab:phylosig-cont)). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmmae \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under a scenario of evolution by BM (\tab \@ref(tab:phylosig-binary)). Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; \tab \@ref(tab:phylosig-binary)). Similar results were obtained using the "strict" dataset (\tab S3).

@@

Community-weighted mean values (CWMs) were smaller for epiphytic than terrestial communities for pinna number, and size-related traits (frond length, frond width, stipe length, and rhizome diameter) (\tval-test, all \pval < 0.05; \fig \@ref(fig:cwm-div)). Standard deviation of community-weighted trait values was larger for size-related traits in terrestrial communities, but smaller for other traits such as frond dissection and pinna number. Specific leaf area decreased with elevation, and was lower for epiphytic relative to terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "sla", "r")`, \pval = `r get_mod_result(model_summ_print, "sla", "p")`). Degree of lamina dissection decreased with elevation in epiphytes, but slightly increased in terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "dissection", "r")`, \pval = `r get_mod_result(model_summ_print, "dissection", "p")`). Frond length and width and rhizome diameter decreased slightly with elevation for terrestrial communities, while increasing slightly for epiphytic communities (\fig \@ref(fig:cwm-div)).

@@

Here, we show for the first time using a broadly sampled phylogenetic dataset that multiple morphological traits related to epiphytic growth are evolutionarily conserved in ferns. We find that epiphtyic ferns are constrained by size, having consistently smaller fronds and rhizomes than terrerstial species. However, contrary to our hypothesis, epiphytic fern communities appear to be more diverse than terrestrial communities. This suggests that environmental filtering plays a strong role in limiting plant size, while allowing for diversification across other trait axes in epiphytes.

>>>>

## Introduction

@@

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA). A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger. Temperature and RH were logged once every \SI{30}{\minute} during two survey periods: a preliminary survey from July 18, 2012 to July 7, 2013 at six plots, and a final survey from July 7, 2013 to July 5, 2014 at 15 plots. Two plots were not included in the environmental survey. We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part of the final survey period (March 12--July 5, 2014), so we used data for the same dates from the preliminary survey period for this datalogger instead. Three other dataloggers also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis. Other days missing data for any datalogger during the final survey period were excluded from the dataset. The final dataset included `r total_days` days of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers. We calculated vapor pressure deficit (VPD) from relative humidity and temperature using the \package{plantecophys} R package [@Duursma2015].

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs \@ref(tab:phylosig-cont), \@ref(tab:phylosig-binary), \fig \@ref(fig:tree)), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \@ref(tab:phylosig-cont)). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \@ref(tab:phylosig-cont)). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmae \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under a scenario of evolution by BM (\tab \@ref(tab:phylosig-binary)). Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; \tab \@ref(tab:phylosig-binary)). Similar results were obtained using the "strict" dataset (\tab S3).

@@

Community-weighted mean values (CWMs) were smaller for epiphytic than terrestrial communities for pinna number, and size-related traits (frond length, frond width, stipe length, and rhizome diameter) (\tval-test, all \pval < 0.05; \fig \@ref(fig:cwm-div)). Standard deviation of community-weighted trait values was larger for size-related traits in terrestrial communities, but smaller for other traits such as frond dissection and pinna number. Specific leaf area decreased with elevation, and was lower for epiphytic relative to terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "sla", "r")`, \pval = `r get_mod_result(model_summ_print, "sla", "p")`). Degree of lamina dissection decreased with elevation in epiphytes, but slightly increased in terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "dissection", "r")`, \pval = `r get_mod_result(model_summ_print, "dissection", "p")`). Frond length and width and rhizome diameter decreased slightly with elevation for terrestrial communities, while increasing slightly for epiphytic communities (\fig \@ref(fig:cwm-div)).

@@

Here, we show for the first time using a broadly sampled phylogenetic dataset that multiple morphological traits related to epiphytic growth are evolutionarily conserved in ferns. We find that epiphytic ferns are constrained by size, having consistently smaller fronds and rhizomes than terrerstial species. However, contrary to our hypothesis, epiphytic fern communities appear to be more diverse than terrestrial communities. This suggests that environmental filtering plays a strong role in limiting plant size, while allowing for diversification across other trait axes in epiphytes.
